---
title: "191003_inclass_EDA"
author: "Gavin Ko"
date: "10/3/2019"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

Group_by is like adding layer on the dataset, be aware.

```{r create weather data}
weather_df = 
  rnoaa::meteo_pull_monitors(c("USW00094728", "USC00519397", "USS0023B17S"),
                      var = c("PRCP", "TMIN", "TMAX"), 
                      date_min = "2017-01-01",
                      date_max = "2017-12-31") %>%
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY", 
                      USC00519397 = "Waikiki_HA",
                      USS0023B17S = "Waterhole_WA"),
    tmin = tmin / 10,
    tmax = tmax / 10,
    month = lubridate::floor_date(date, unit = "month")) %>%
  select(name, id, date, month, everything())
```

```{r first groupby function}
weather_df %>%
  group_by(name, month)
```
There's nothing new, or is there? The group index is on top of the dataset.
When you put weather_df itself, it doesn't show the group_by layer.

```{r groupby and observation}
weather_df %>% 
  group_by(month) %>% 
  summarize(
    n_obs = n(), #create new variable to accomodate existing values
    n_unique = n_distinct(date)) #check distinct value
```

You don't need to do `group_by` and `summarize` every single time. 
Let's use count.

```{r count function}
weather_df %>% 
  count(name, month, name = "n_days") 
#bad example of using name: it's a parameter to assign value column names.
```

How do you generate table?
```{r Bad table}
## don't ever do this
## weather_df %>% 
## pull(name) %>% 
## table() 
## Base R functions are not allowed.
```
```{r How do we really generate table}
weather_df %>% 
  count(name) %>% 
  knitr::kable()
```

Created 2 x 2 table...
```{r 2 x 2 table}
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when( #just as ifelse, just choose what you like.
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE ~ ""
    )
  ) %>% 
  group_by(name, cold) %>% 
  count() %>% 
  pivot_wider(
    names_from = name,
    values_from = cold
  )
```
but it's 4 x 2, what are you taking about?

```{r better way to do 2 x 2 table: tabyl}
weather_df %>% 
  filter(name != "Waikiki_HA") %>% 
  mutate(
    cold = case_when( 
      tmax < 5 ~ "cold",
      tmax >= 5 ~ "not cold",
      TRUE ~ ""
    )
  ) %>% 
  janitor::tabyl(name, cold) #tabyl package
```
Special note: Jeff is AGAINST 2 x 2 table.

Let's move on to general summaries. 

```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    n = n(),
    mean_tmax = mean(tmax, na.rm = T),
    sd_tmax = sd(tmax, na.rm = T),
    median_prcp = median(tmax, na.rm = T)
 ) %>% 
  ggplot(aes (x = month, y = mean_tmax, color = name) ) + geom_point() + geom_line()
# Waikiki has NA value because of 3 rows of missing data.
# You cannot take mean of sth doesn't exist.
# Double check before na.rm, see how did the NA's come from
# multiple condition is also allowed
```

This is too complicated. Let's make it more simple.
```{r}
weather_df %>% 
  group_by(name, month) %>% 
  summarize(
    n = n(),
    mean_tmax = mean(tmax, na.rm = T),
 ) %>% 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) %>% 
  knitr::kable(digits = 1) #rounding
```

Don't forget you can ungroup by using `ungroup()`!

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    mean_tmax = mean(tmax, na.rm = T), # specific to all value within each group
    centered_tmax = tmax - mean_tmax
  ) %>% 
  ggplot(aes (x = date, y = centered_tmax, color = name)) + geom_point()

```

Window functions in group mutates, talking about rankings.

```{r comparing the coldest days}
weather_df %>% 
  group_by(name, month) %>% 
  mutate(
    tmax_rank = min_rank(tmax) #min rank indicates the smaller comes in the front rank.
  ) %>% 
  filter(tmax_rank == 1)
```
For more information, check `?min_rank` for full set of variables.

Next, let's check lags and leads.
When we want to check the day by day differences, which location fluctuate the most, ...

```{r}
weather_df %>% 
  group_by(name) %>% 
  mutate(
    lagged_tmax = lag(tmax), #copy all the information, shift by one row down
    one_day_tmax_change = tmax - lagged_tmax,
    leaded_tmax = lead(tmax) #the opposite of lagging
  ) %>% 
  summarize(
    sd_daily_change = sd(one_day_tmax_change, na.rm = T)
  )
# don't forget, 'summarize' is not 'summary'
```





